DATA 621---Business Analytics and Data Mining

### 2020-09-18 HW1/HW1_draft_kellogg.Rmd
Significant rewrite, especially of the conclusion, in light of model performance. Added TOC. Spelling and Grammar reviewed.

---
title: 'DATA 621---Business Analytics and Data Mining'
subtitle: 'Fall 2020---Group 2---Homework #1'
author: Avraham Adler, Samantha Deokinanan, Amber Ferger, John Kellogg,
    Bryan Persaud, Jeff Shamp
date: "9/27/2020"
output:
  pdf_document: default
  html_document: default
---

@@

  * **TEAM_PITCHING_SO**: The largest outlier is close to 20,000, which averages
  ~123 strikeouts per game. This is only possible if every one of the 162 games
  went approximately 40 innings, and each out was a strikeout---which clearly has
  never happened.
  * **TEAM_PITCHING_H**: The largest outlier for this vaiable is over 30,000
  which is simlarly highly unlikely.

@@

# BUILD MODELS
## Base Model
We will start with a simple linear model to serve as a baseline. This includes
all variables in the dataset.

@@

## Model 1
Model 1 includes all variables except Singles, Triples, Base Hits, Strikeouts by
pitchers, Walks allowed, and Batters Hit by Pitch. The final variables were
chosen as a result of backwards `dplyr::selection` based on null hypothesis
testing for non-zero slope.

@@

A fully specified model with all the interactions will be the starting point for
the stepwise regression, which uses AIC as its target metric.

```{r model2}
crazyModel <- lm(TARGET_WINS ~
  (TEAM_BASERUN_CS + TEAM_BASERUN_SB) ^ 2 + (TEAM_BATTING_2B + TEAM_BATTING_3B +
  TEAM_BATTING_BB + TEAM_BATTING_H + TEAM_BATTING_HR + TEAM_BATTING_SO) ^ 2 +
  (TEAM_FIELDING_DP + TEAM_FIELDING_E) ^ 2 + (TEAM_PITCHING_BB +
  TEAM_PITCHING_H + TEAM_PITCHING_HR + TEAM_PITCHING_SO) ^ 2, data = ts2)
stepModel <- stepAIC(crazyModel, scope = list(upper = ~
  (TEAM_BASERUN_CS + TEAM_BASERUN_SB) ^ 2 + (TEAM_BATTING_2B + TEAM_BATTING_3B +
  TEAM_BATTING_BB + TEAM_BATTING_H + TEAM_BATTING_HR + TEAM_BATTING_SO) ^ 2 +
  (TEAM_FIELDING_DP + TEAM_FIELDING_E) ^ 2 + (TEAM_PITCHING_BB +
  TEAM_PITCHING_H + TEAM_PITCHING_HR + TEAM_PITCHING_SO) ^ 2, lower = ~ 1),
  direction = 'both', trace = 0)
sstm <- summary(stepModel)
stmr2 <- sstm$r.squared
stmar2 <- sstm$adj.r.squared
stmAIC <- prettyNum(AIC(stepModel), digits = 2L, big.mark = ',')
stmRMSE <- sqrt(mean(stepModel$residuals ^ 2))
stmF <- prettyNum(sstm$fstatistic, digits = 5L)
stmFp <- pf(sstm$fstatistic[[1]], sstm$fstatistic[[2]], sstm$fstatistic[[3]],
            lower.tail = FALSE)
```

### Coefficient Discussion:
```{r model2_discuss}
knitr::kable(sstm$coefficients, digits = 3L,
             caption = 'Step Model with Pairwise Class Interactions Output')
```

Most of the coefficients are reasonable within the context of the game of
baseball. These two statements are axiomatic:

@@

The curiosities above can *possibly* be resolved by looking at the interaction
terms. The pitching and batting home run and strikeout coefficients are an order
of magnitude greater than all the others. It is likely that the negative
coefficients in some of the batting stats and positive coefficents in some of
the pitching stats are being used to "temper" the effect of SOs and HRs.

@@

### Model 3
The final model is a Stepwise Regression with Repeated k-fold Cross-Validation,
and higher order polynomials variables were introduced into the full model.

@@

```{r model3}
stepdata = subset(train_set, select = -c(AB, SLG, TEAM_BATTING_HBP))
stepdata = cbind(stepdata, sapply(stepdata[2:length(stepdata)], function(x) x^2))
names(stepdata) = make.unique(names(stepdata), sep = "_")
model = lm(TARGET_WINS ~ ., data = stepdata)
cooksd = cooks.distance(model)
influential = as.numeric(names(cooksd)[(cooksd > 4*mean(cooksd, na.rm = TRUE))])
stepdata.2 = stepdata[-c(influential), ]
# Set up repeated k-fold cross-validation
set.seed(525)
train.control = trainControl(method = "cv", number = 10)
# Train the model
step.model = train(TARGET_WINS ~ ., data = stepdata.2, method = "lmStepAIC", 
                   trControl = train.control, trace = FALSE)
# Final model 
step.model = lm(TARGET_WINS ~ TEAM_BATTING_H + TEAM_BATTING_2B + TEAM_BASERUN_SB + 
                  TEAM_BASERUN_CS + TEAM_PITCHING_H + TEAM_PITCHING_HR + TEAM_FIELDING_E + 
                  TEAM_FIELDING_DP + TEAM_BATTING_BB_1 + TEAM_BATTING_SO_1 + 
                  TEAM_BASERUN_CS_1 + TEAM_PITCHING_H_1 + TEAM_PITCHING_BB_1 + 
                  TEAM_FIELDING_E_1 + TEAM_FIELDING_DP_1 + TEAM_BATTING_1B_1, 
                data = stepdata.2)
# Summary of the model
mod_sum3 = summary(step.model)
```

**Coefficient Discussion: **
```{r model3_discuss}
mod_sum3[["coefficients"]] %>% 
  knitr::kable(digits = 3L, 
               caption = 'K-fold Step Model with Higher Order Polynomials Output')
```

@@

## SELECTING A MODEL & PREDICTIONS

In order to select the best model to make predictions, we looked at some measurements that tells us how well each model fits the training. These include the 1) $R^2$, which represents the proportion of the variance explained by a model; 2) $adj R^2$, which is a modified version of $R^2$; 3) *Root Mean Squared Error* (RMSE), which is the square root of the mean squared error; and 4) *Akaike Information Criterion* (AIC), which is an estimator of out-of-sample prediction error.

```{r performance}
# Model accuracy
results = data.frame(
  Models = c("Model #1: Backwards" , "Model #2: Pairwise", "Model #3: K-fold"),
  R.squared = c(mod_sum1[["r.squared"]], stmr2, mod_sum3[["r.squared"]]), 
  adj.R.squared = c(mod_sum1[["adj.r.squared"]], stmar2, mod_sum3[["adj.r.squared"]]),
  RSME = c(sqrt(mean(lm_1$residuals ^ 2)), stmRMSE, sqrt(mean(step.model$residuals ^ 2))),
  AIC = c(AIC(lm_1), AIC(stepModel), AIC(step.model)))
results %>% knitr::kable(digits = 3L, 
                         caption = 'Performance Statistics & Error Measure of Models')
```

It was an interesting process when it came comparing the three models and selecting which would be our best model given that their performance statistics and error measurements were not significantly different from each other. From the table above, it is apparent that Model #3: K-fold accounts for nearly 45% of the variation in the dependent variable with the independent variables because $R^2 = 0.456$, $adjR^2 = 0.450$ which is acceptable as a good model. However, Model #1: Backwards has a smallest RSME of all the models, which ranks as number one as the criteria we are using to decide on a model. This measure suggest how far off an actual value is from the model's prediction for that value, and it indicates that there is a greater reduction in the randomness for Model #1: Backwards than the other model. But, how different is the RSME of Model #1: Backwards versus the RSME of Model #3: K-fold? We agreed that it is insignificant. Therefore, because the RMSE and adjusted $R^2$ statistics already include a minor adjustment for the number of coefficients estimated, to evaluate the model complexity, we compared the AIC. As a result, Model #1: Backwards would have been the way to go, however, it was deemed that the AIC difference with Model #3 is also insignificant.

### PREDICTIONS

Using the evaluation data set and the three models, a comparison in the predictions statistic with the training data was conducted. Interestingly, while Model #1: Backwards and Model #3: K-fold resulted in better performance statistics and error measurements, it is evident that Model #2: Pairwise yields closer predictions when the other two slightly under-predicts. With all in agreement, under-predicting wins and leaving teams and managers with potential perspective of costs is worse than over-predicting and causing inconvenience.

```{r eval_set}
eval_set = eval_set %>% na.omit() %>%
  summarise(iqr = IQR(TEAM_PITCHING_SO)) %>%
  bind_cols(eval_set) %>%
  filter(TEAM_PITCHING_SO > quantile(TEAM_PITCHING_SO, probs = c(0.25), na.rm = TRUE) 
         - 1.5 * iqr, 
         TEAM_PITCHING_SO < quantile(TEAM_PITCHING_SO, probs = c(0.75), na.rm = TRUE) 
         + 1.5 * iqr) %>%
  dplyr::select(-iqr)

@@

```{r predictions}
predictions = data.frame(predict(lm_1, newdata = test_df1), 
                         predict(stepModel, test_df2), 
                         predict(step.model, newdata = test_df3))
results = cbind(
  Datasets = c('Training Data', 'Model #1: Backwards', 'Model #2: Pairwise',
               'Model #3: K-fold'), 
  rbind(psych::describe(train_set$TARGET_WINS), psych::describe(predictions))) 
rownames(results) = NULL
results[,-c(2,8,11)] %>% 
  knitr::kable(digits = 2L, caption = 'Prediction Comparison on Evaluation Set')
```

Therefore, based on these reasoning, it is unanimous that **Model #2: Pairwise** is our final model because the coefficients of the model with it's the interacting variables makes for better, realistic predictions.

## CONCLUSIONS

## CODE APPENDIX
The code chunks below represent the R code called in order during the analysis.
They are reproduced in the appendix for review and comment.
```{r appendix, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```
```{r setup}
```
```{r loadData}
```
```{r wins}
```
```{r pctNull}
```
```{r negFactors}
```
```{r posFactors}
```
```{r summaryInfo}
```
```{r fewOutliers}
```
```{r manyOutliers}
```
```{r otherOutliers}
```
```{r corrPlot}
```
```{r outliers_1}
```
```{r iqr_removal_SO}
```
```{r iqr_removal_H}
```
```{r outlier_boxplot}
```
```{r impute_train}
```
```{r singles}
```
```{r make_slg_obp}
```
```{r slugging_compare}
```
```{r baseModel}
```
```{r cooks_distance}
```
```{r model2C+S}
```
```{r model2}
```
```{r model3}
```
```{r performance}
```
```{r eval_set}
```
```{r predictions}
```

>>>>

---
title: 'DATA 621---Business Analytics and Data Mining'
subtitle: 'Fall 2020---Group 2---Homework #1'
author: Avraham Adler, Samantha Deokinanan, Amber Ferger, John Kellogg,
    Bryan Persaud, Jeff Shamp
date: "9/27/2020"
output:
  pdf_document:
    toc: TRUE
    toc_depth: 4
  html_document: default
---

@@

  * **TEAM_PITCHING_SO**: The largest outlier is close to 20,000, which averages
  ~123 strikeouts per game. This is only possible if every one of the 162 games
  went approximately 40 innings, and each out was a strikeout---which clearly has
  never happened.
  * **TEAM_PITCHING_H**: The largest outlier for this variable is over 30,000
  which is similarly highly unlikely.

@@

# MODEL TRAINING
## Model 1
### Base Model
We will start with a simple linear model to serve as a baseline. This includes
all variables in the dataset.

@@

### Enhanced Model
Model 1 includes all variables except Singles, Triples, Base Hits, Strikeouts by
pitchers, Walks allowed, and Batters Hit by Pitch. The final variables were
chosen as a result of backwards `dplyr::selection` based on null hypothesis
testing for non-zero slope.

@@

### Stepwise Model
A fully specified model with all the interactions will be the starting point for
the stepwise regression, which uses AIC as its target metric. For interest, this
model will be compared with both the null model and the fully-specified model
from where the stepwise process begins

```{r model2}
nullModel <- lm(TARGET_WINS ~ 1, data = ts2)
fullModel <- lm(TARGET_WINS ~
  (TEAM_BASERUN_CS + TEAM_BASERUN_SB) ^ 2 + (TEAM_BATTING_2B + TEAM_BATTING_3B +
  TEAM_BATTING_BB + TEAM_BATTING_H + TEAM_BATTING_HR + TEAM_BATTING_SO) ^ 2 +
  (TEAM_FIELDING_DP + TEAM_FIELDING_E) ^ 2 + (TEAM_PITCHING_BB +
  TEAM_PITCHING_H + TEAM_PITCHING_HR + TEAM_PITCHING_SO) ^ 2, data = ts2)
stepModel <- stepAIC(fullModel, scope = list(upper = ~
  (TEAM_BASERUN_CS + TEAM_BASERUN_SB) ^ 2 + (TEAM_BATTING_2B + TEAM_BATTING_3B +
  TEAM_BATTING_BB + TEAM_BATTING_H + TEAM_BATTING_HR + TEAM_BATTING_SO) ^ 2 +
  (TEAM_FIELDING_DP + TEAM_FIELDING_E) ^ 2 + (TEAM_PITCHING_BB +
  TEAM_PITCHING_H + TEAM_PITCHING_HR + TEAM_PITCHING_SO) ^ 2, lower = ~ 1),
  direction = 'both', trace = 0)
snm <- summary(nullModel)
sfm <- summary(fullModel)
sstm <- summary(stepModel)
stmr2 <- sstm$r.squared
stmar2 <- sstm$adj.r.squared
stmAIC <- AIC(stepModel)
stmRMSE <- sqrt(mean(stepModel$residuals ^ 2))
```

### Discussion
#### Models
```{r model2_discuss}
knitr::kable(snm$coefficients, digits = 3L,
             caption = 'Null Model')
knitr::kable(sfm$coefficients, digits = 3L,
             caption = 'Full Model with All Features & Pairwise Class Interactions')
knitr::kable(sstm$coefficients, digits = 3L,
             caption = 'Stepwise Model with Pairwise Class Interactions')
```

The null model expects teams to lose slightly more than half their games. The
full model has a lower intercept, which indicates that the impact of all the
features explains why teams *win* slightly more than just the grand mean does.
The stepwise model is similar, having a lower intercept than the null but a
higher intercept that the full model.

Also, the full model has the higher \(R^2\) and the stepwise model has the
higher *adjusted* \(R^2\). This is expected, as there is no penalty for adding
more parameters in regular \(R^2\) measures.

The **stepwise** model will be the one discussed below.

#### Coefficients
Most of the coefficients are reasonable within the context of the game of
baseball. These two statements are axiomatic:

@@

The curiosities above can *possibly* be resolved by looking at the interaction
terms. The pitching and batting home run and strikeout coefficients are an order
of magnitude greater than all the others. It is likely that the negative
coefficients in some of the batting stats and positive coefficients in some of
the pitching stats are being used to "temper" the effect of SOs and HRs.

@@

## Model 3
The final model is a Stepwise Regression with Repeated k-fold Cross-Validation,
and higher order polynomials variables were introduced into the full model.

@@

```{r model3}
stepdata <- subset(train_set, select = -c(AB, SLG, TEAM_BATTING_HBP))
stepdata <- cbind(stepdata, sapply(stepdata[2:length(stepdata)], function(x) x^2))
names(stepdata) <- make.unique(names(stepdata), sep = "_")
model <- lm(TARGET_WINS ~ ., data = stepdata)
cooksd <- cooks.distance(model)
influential <- as.numeric(names(cooksd)[(cooksd > 4*mean(cooksd, na.rm = TRUE))])
stepdata.2 <- stepdata[-c(influential), ]
# Set up repeated k-fold cross-validation
set.seed(525)
train.control <- trainControl(method = "cv", number = 10)
# Train the model
step.model <- train(TARGET_WINS ~ ., data = stepdata.2, method = "lmStepAIC", 
                   trControl = train.control, trace = FALSE)
# Final model 
step.model <- lm(TARGET_WINS ~ TEAM_BATTING_H + TEAM_BATTING_2B + TEAM_BASERUN_SB + 
                  TEAM_BASERUN_CS + TEAM_PITCHING_H + TEAM_PITCHING_HR + TEAM_FIELDING_E + 
                  TEAM_FIELDING_DP + TEAM_BATTING_BB_1 + TEAM_BATTING_SO_1 + 
                  TEAM_BASERUN_CS_1 + TEAM_PITCHING_H_1 + TEAM_PITCHING_BB_1 + 
                  TEAM_FIELDING_E_1 + TEAM_FIELDING_DP_1 + TEAM_BATTING_1B_1, 
                data = stepdata.2)
# Summary of the model
mod_sum3 <- summary(step.model)
```

### Coefficient Discussion
```{r model3_discuss}
mod_sum3[["coefficients"]] %>% 
  knitr::kable(digits = 3L, 
               caption = 'K-fold Step Model with Higher Order Polynomials Output')
```

@@

# MODEL SELECTION
## Criteria
In order to select the best model to make predictions, we looked at the
following goodness of fit metrics:

 1. \(R^2\), which represents the proportion of the variance explained by the
 model
 2. Adjusted \(R^2\), which is a modified version of \(R^2\) taking parsimony
 into consideration
 3. *Root Mean Squared Error* (RMSE), which is the square root of the mean
 squared difference between the observation and the fitted value
 4. *Akaike Information Criterion* (AIC), which is an information-criteria based
 estimate of the Kullback-Leibler divergence from the current model to the
 "true" model. One of the nice results of the theory of information-criteria
 based measures is that there is included recognition of parsimony.

## Performance
```{r performance}
# Model accuracy
results <- data.frame(
  Models = c("Model #1: Backwards" , "Model #2: Pairwise", "Model #3: K-fold"),
  R.squared = c(mod_sum1[["r.squared"]], stmr2, mod_sum3[["r.squared"]]), 
  adj.R.squared = c(mod_sum1[["adj.r.squared"]], stmar2,
                    mod_sum3[["adj.r.squared"]]),
  RSME = c(sqrt(mean(lm_1$residuals ^ 2)), stmRMSE,
           sqrt(mean(step.model$residuals ^ 2))),
  AIC = c(AIC(lm_1), AIC(stepModel), AIC(step.model)))
results %>% knitr::kable(digits = 3L, 
                         caption = 'Performance Statistics & Error Measure of Models')
```

It was interesting to compare the three models given that their performance
statistics and error measurements were not significantly different from each other.
**Model #2: Pairwise** was the worst performing model for all of our metrics, and
was eliminated immediately. **Model #3: K-fold** accounts for nearly 45% of the
variation in the dependent variable with the independent variables because its
\(R^2\) is 0.456 and its adjusted \(R^2\) is = 0.450, both of which are
acceptable for a good model. However, **Model #1: Backwards** has the smallest
RSME of all the models. RMSE is a measure of how far off predictions are from
the actual, giving extra weight to outliers through summing the squared
differences. This demonstrates a greater reduction in the randomness by
**Model #1: Backwards**. But its difference from **Model #3: K-fold** is only
0.022. It is hard to judge if that is a significant difference. We thus also
considered AIC. With AIC, it is not the magnitude of the value which is
important, but the difference between the various model AIC's. The magnitude is
a function of the number of observations. Accepted rules of thumb in the
statistical literature are that if the difference is greater than 10, there is
effectively no support for the model with the larger AIC
(Section 2.6, Burnham & Anderson 2002). As a result, **Model #1: Backwards**
is our selected model.

## Discussion
This is an interesting model in that given no other information, a team would be
expected to win negative games, which is clearly impossible. That being said, it
is an indication of the predictive power of the engineereed **slugging** feature
that allows small increments in slugging to result in significant shifts in
target wins. This engineered feature contains within it numerous interactions.
It combines scaled versions of some features and normalizes them for the
combination of other features. Perhaps that is why it is so significant.

# PREDICTIONS
**Model #1: Backwards** was used to predict the target wins base on the
evaluation set, once processed similarly to the training data. As the number of
games won must be integral, the values were rounded to the closest integer.

```{r eval_set, eval=FALSE}
eval_set = eval_set %>% na.omit() %>%
  summarise(iqr = IQR(TEAM_PITCHING_SO)) %>%
  bind_cols(eval_set) %>%
  filter(TEAM_PITCHING_SO > quantile(TEAM_PITCHING_SO, probs = c(0.25), na.rm = TRUE) 
         - 1.5 * iqr, 
         TEAM_PITCHING_SO < quantile(TEAM_PITCHING_SO, probs = c(0.75), na.rm = TRUE) 
         + 1.5 * iqr) %>%
  dplyr::select(-iqr)

@@

```{r predictions}
eval_set <- paste0(urlRemote, pathGithub, fileEval) %>% read.csv()
eval_set <- eval_set %>% na.omit() %>%
  summarise(iqr = IQR(TEAM_PITCHING_SO)) %>%
  bind_cols(eval_set) %>%
  filter(TEAM_PITCHING_SO > quantile(TEAM_PITCHING_SO, probs = c(0.25), na.rm = TRUE) 
         - 1.5 * iqr, 
         TEAM_PITCHING_SO < quantile(TEAM_PITCHING_SO, probs = c(0.75), na.rm = TRUE) 
         + 1.5 * iqr) %>%
  dplyr::select(-iqr)

eval_set <- eval_set %>% na.omit() %>%
  summarise(iqr = IQR(TEAM_PITCHING_H)) %>%
  bind_cols(eval_set) %>%
  filter(TEAM_PITCHING_H > quantile(TEAM_PITCHING_H, probs = c(0.25), na.rm = TRUE) 
         - 1.5 * iqr,
         TEAM_PITCHING_H < quantile(TEAM_PITCHING_H, probs = c(0.75), na.rm = TRUE) 
         + 1.5 * iqr) %>%
  dplyr::select(-iqr)

eval_set <- complete(mice(data = eval_set, method = "pmm", seed = 9450, print=FALSE), 3)

eval_set <- eval_set %>% dplyr::select(-INDEX) %>%
  mutate_if(is.integer, as.numeric) %>%
  mutate(TEAM_BATTING_1B = TEAM_BATTING_H - (TEAM_BATTING_2B + TEAM_BATTING_3B +
                                               TEAM_BATTING_HR))
eval_set <- eval_set %>%
  mutate(AB = TEAM_BATTING_SO + TEAM_BATTING_BB + TEAM_BATTING_H,
         SLG = (TEAM_BATTING_1B + 2 * TEAM_BATTING_2B + 3 * TEAM_BATTING_3B +
                  4 * TEAM_BATTING_HR) / AB)
predictions <- round(predict(lm_1, newdata = eval_set))
summary(predictions)
predictions
```

# REFERENCES

 * Burnham, Kenneth P., and David R. Anderson. 2002. Model Selection and
 Multimodel Inference: A Practical Information-Theoretic Approach. Second.
 New York: Springer Science+Business Media, Inc.

# CODE APPENDIX
The code chunks below represent the R code called in order during the analysis.
They are reproduced in the appendix for review and comment.
```{r appendix, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```
```{r setup}
```
```{r loadData}
```
```{r wins}
```
```{r pctNull}
```
```{r negFactors}
```
```{r posFactors}
```
```{r summaryInfo}
```
```{r fewOutliers}
```
```{r manyOutliers}
```
```{r otherOutliers}
```
```{r corrPlot}
```
```{r outliers_1}
```
```{r iqr_removal_SO}
```
```{r iqr_removal_H}
```
```{r outlier_boxplot}
```
```{r impute_train}
```
```{r singles}
```
```{r make_slg_obp}
```
```{r slugging_compare}
```
```{r baseModel}
```
```{r cooks_distance}
```
```{r model2C+S}
```
```{r model2}
```
```{r model3}
```
```{r performance}
```
<!--```{r eval_set}
```-->
```{r predictions}
```
